***De novo* clustering methods out-perform reference-based methods for assigning 16S rRNA gene sequences to operational taxonomic units**

### 2015-11-13 ResponseToReviewers.md

---
title: "***Response to reviewers**"
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: peerj.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

> - The findings on "problems with reference-based clustering in general and as implemented in QIIME" are striking and, indeed, slightly disturbing. To be honest, I would not have expected such strong effects of sequence order for mappings against the QIIME reference database. I believe that these results will contribute to more debates in the field, and ultimately lead to further long-needed benchmarks on standard operating procedures for widely used software.

Thank you for highlighting these three points.

> - p.12, ll.260-1. This statement is too harsh based on the given data. The SWARM algorithm is designed to not work with "hard" thresholds in the classical sense, so it is not surprising that in the given testing framework, different "thresholds" would appear optimal for differentially deeply sequenced datasets.

We attempted to soften this language. Regardless, in our opinion, the inability to define a "hard" threshold makes it difficult to compare results across studies.

### 2015-11-13 Schloss_Cluster_PeerJ_2015.Rmd
Editing

**Methods.** Our analysis implemented five *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and the open and closed-reference methods. Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.

@@

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014]. Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011Assessing]. For example, all sequences assigned to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches; however, it would still suffer from the problems associated with database quality and classification error.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014]. These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments. Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @White2010]. This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010]. A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014]. Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities that lack the phylogenetic diversity of real microbial communities [@Kozich2013; @Schloss2011Reducing]. Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014]. This is problematic because it does not solve the fundamental question of which method is most correct. The concept of ecological consistency as a metric of quality asserts that sequences that cluster into the same OTU should share similar ecological affiliations [@Koeppel2013; @Schmidt2014Ecological; @Preheim2013]. Although this is an intriguing approach and is a metric that can be quantified, it is unclear how the metric would be validated. We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011Assessing]. We were able to synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975]. MCC is can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0. The strength of the MCC, as implemented by Schloss et al. [-@Schloss2011Assessing], is that it is an objective approach to assessing the quality of the OTU assignments that can be calculated for any set of OTU assignments where there is a distance matrix and a specific threshold without relying on an external reference.

A recent analysis by He and colleagues [-@He2015] characterized the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset. Their concept of stability did not account for the accuracy of the OTU assignments and instead focused on the precision of the assignments. A method may be very precise, but low in accuracy. In the current analysis, we assessed the accuracy and precision of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of accuracy. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset. The He study used the MCC to quantify the degree to which pairs of sequences were in the same OTUs in subsampled and full datasets. A more robust alternative, which used metrics that quantify the mutual information held between two sets of clusterings, has been applied to assess inter-method variation [@Schmidt2014Limits]. To maintain consistency with the original He study, we also calculated the MCC value as they described. The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset. Among the *de novo* methods they observed that the abundance-based greedy clustering (AGC) method generated the most stable OTUs followed by the single linkage (SL), distance-based greedy clustering (DGC), complete linkage (CL), and average linkage (AL) methods. We first calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (see Methods for description; Figure 1A). Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times. Multiple subsamplings were necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them. Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception. The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data. This result appeared out of character and was not explained by the authors. They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`). This result indicates that the AL assignments were far more stable than indicated in the He study. Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.

Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a portion of the dataset did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset. The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data. In addition, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data. Our results support those of these previous studies (Figure 2). It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data. The number of OTUs observed was largest using the CL method, followed by the open-reference method. The AL, AGC, and DGC methods all provided comparable numbers of OTUs. Finally, the closed-reference and SL methods generated the fewest number of OTUs.

Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs [@Anderson2001]. Although they were able to detect significant p-values, they appeared to be of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis was tautological since, by definition, representing communities based on their unstable OTUs would yield differences. Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets. It is not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method. This entire analysis assumed that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.

@@

***How do the various methods vary in the quality of their OTU assignments?*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the MCC. We used the average MCC value as a measure of a method's quality and its variation as a measure of its consistency. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the quality of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produce OTU assignments that are equally good. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs. Our analysis indicates that the most optimal method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.

@@

***How generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing]. The large number of sequences that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010]. With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012Stabilization; @Kozich2013]. The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013]. Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data, the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent.

Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations. Therefore, although the methods have a stochastic component, the OTU assignments do not vary meaningfully between runs. The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the the overall quality of the assignments were consistent. Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods were consistently better than either of the reference-based methods.

***How does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014]. Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community. Similar to the authors of the He study, the Swarm developers suggest that methods are needed that are insensitive to input order. Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A). When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D). Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D). Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively. This suggests that distance-based OTU definitions are not consistent across datasets when using the Swarm algorithm, although they do appear to be within the neighborhood of 3%. Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets. Our ability to apply AL to the large mouse dataset and even larger datasets suggests [e.g. @Schubert2015; @Zackular2015] that it is not necessary to sacrifice OTU assignment quality for speed.

@@

We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations. When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`). If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`). To better understand this result, we further analyzed QIIME's reference database. We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other. When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization. Among the `r n_v4_overlap` reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene. When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences. Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other. The presence of duplicate V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method. We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other as a way to overcome the limitations of the USEARCH search heuristic.

>>>>

**Methods.** Our analysis implemented six *de novo* clustering algorithms including the single linkage, complete linkage, average linkage, abundance-based greedy clustering, distance-based greedy clustering, and Swarm and the open and closed-reference methods. Using two previously published datasets we used the Matthew’s Correlation Coefficient (MCC) to assess the stability and quality of OTU assignments.

@@

The third approach, open-reference clustering, is a hybrid of the closed-reference and *de novo* approaches [@NavasMolina2013; @Rideout2014]. Open-reference clustering involves performing closed-reference clustering followed by *de novo* clustering on those sequences that are not sufficiently similar to the reference. In theory, this method should exploit the strengths of both closed-reference and *de novo* clustering; however, the different OTU definitions employed by commonly used closed-reference and *de novo* clustering implementations pose a possible problem when the methods are combined. An alternative to this approach has been to classify sequences to a bacterial family or genus and then assign those sequences to OTUs within those taxonomic groups using the average linkage method [@Schloss2011Assessing]. For example, all sequences classified as belonging to the *Porphyromonadaceae* would then be assigned to OTUs using the average linkage method using a 3% distance threshold. Those sequences that did not classify to a known family would also be clustered using the average linkage method. An advantage of this approach is that it lends itself nicely to parallelization since each taxonomic group is seen as being independent and can be processed separately. Such an approach would overcome the difficulty of mixing OTU definitions between the closed-reference and *de novo* approaches; however, it would still suffer from the problems associated with database quality and classification error.

The growth in options for assigning sequences using each of these three broad approaches has been considerable. It has been difficult to objectively assess the quality of OTU assignments. Some have focused on the time and memory required to process a dataset [@Sun2009; @Cai2011; @Rideout2014; @Mah2014]. These are valid parameters to assess when judging a clustering method, but have little to say about the quality of the OTU assignments. Others have attempted to judge the quality of a method by its ability to generate data that parallels classification data [@Cai2011; @Sun2011; @White2010]. This approach is problematic because bacterial taxonomy often reflects historical biases amongst bacterial systematicists. Furthermore, it is well known that the rates of evolution across lineages are not the same [@Wang2007; @Schloss2010]. A related approach has used clustering of mock community data to evaluate methods [@Edgar2013; @Mah2014; @Barriuso2011; @Bonder2012; @Chen2013; @Huse2010; @May2014]. Yet these approaches ignore the effects of sequencing errors that tend to accumulate with sequencing depth and represent highly idealized communities that lack the phylogenetic diversity of real microbial communities [@Kozich2013; @Schloss2011Reducing]. Others have assessed the quality of clustering based on their ability to generate the same OTUs generated by other methods [@Schmidt2014Limits; @Rideout2014]. This is problematic because it does not solve the fundamental question of which method is optimal. The concept of ecological consistency as a metric of quality asserts that sequences that cluster into the same OTU should share similar ecological affiliations [@Koeppel2013; @Schmidt2014Ecological; @Preheim2013]. Although this is an intriguing approach and is a quantitative metric, it is unclear how the metric would be objectively validated. We recently proposed an approach for evaluating OTU assignments using the distances between pairs of sequences [@Schloss2011Assessing]. We were able to synthesize the relationship between OTU assignments and the distances between sequences using the Matthew's correlation coefficient [MCC; @Matthews1975]. MCC is can be interpreted as representing the correlation between the observed and expected classifications and can vary between -1.0 and 1.0. The strength of the MCC, as implemented by Schloss et al. [-@Schloss2011Assessing], is that it is an objective approach to assessing the quality of the OTU assignments that can be calculated for any set of OTU assignments where there is a distance matrix and a specific threshold without relying on an external reference.

A recent analysis by He and colleagues [-@He2015] characterized the three general clustering approaches by focusing on what they called stability. They defined stability as the ability of a method to provide the same clustering on a subset of the data as was found in the full dataset. Their concept of stability did not account for the quality of the OTU assignments and instead focused on the precision of the assignments. A method may be very stable, but of poor quality. In the current analysis, we assessed the quality and stability of the various clustering methods. Building on our previous analysis of clustering methods, our hypothesis was that the methods praised by the He study for their stability actually suffered a lack of quality. In addition, we assess these parameters in light of sequence quality using the original 454 dataset and a larger and more modern dataset generated using the MiSeq platform.

@@

First, we sought to quantify whether the OTU assignments observed for a subset of the data represented the same assignments that were found with the full dataset. The He study used the MCC to quantify the degree to which pairs of sequences were in the same OTUs in subsampled and full datasets. A more robust approach would utilize metrics that quantify the mutual information held between two sets of clusterings and has been applied to assess inter-method variation in OTU composition [@Schmidt2014Limits]. To maintain consistency with the original He study, we also calculated the MCC value as they described. The He study found that when they used the open and closed-reference methods the OTUs formed using the subsetted data most closely resembled those of the full dataset. Among the *de novo* methods they observed that the AGC method generated the most stable OTUs followed by the single linkage (SL), DGC, complete linkage (CL), and average linkage (AL) methods. We first calculated the MCC for the OTU assignments generated by each of the clustering methods using 20, 40, 60, and 80% of the sequences relative to the OTU composition formed by the methods using the full dataset (see Methods for description; Figure 1A). Similar to the He study, we replicated each method and subsampled to the desired fraction of the dataset 30 times. Multiple subsamplings were necessary because a random number generator is used in some of the methods to break ties where pairs of sequences have the same distance between them. Across these sequencing depths, we observed that the stability of the OTUs generated by the SL and CL methods were highly sensitive to sampling effort relative to the OTUs generated by the AL, AGC, and DGC methods (Figure 1A). Our results (Figure 1B) largely confirmed those of Figure 4C in the He study with one notable exception. The He study observed a broad range of MCC values among their AL replicates when analyzing OTUs generated using 60% of the data. This result appeared out of character and was not explained by the authors. They observed a mean MCC value of approximately 0.63 (95% confidence interval between approximately 0.15 and 0.75). In contrast, we observed a mean value of `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "mean"], 2)` (95% confidence interval between `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "lci"], 2)` and `r round(he_stability[he_stability$method=="an" & he_stability$fraction==0.6, "uci"], 2)`). This result indicates that the AL assignments were far more stable than indicated in the He study. Regardless, although the assignments are quite stable, it does support the assertion that the OTU assignments observed for the subset of the data do not perfectly match the assignments that were found with the full dataset as they did with the reference-based methods; however, the significance of these differences is unclear.

Second, the He study and the original Roesch study showed that rarefaction curves calculated using CL-generated OTU assignments obtained using a subsample of the sequencing data did not overlap with rarefaction curves generated using OTU assignments generated from the full dataset. The He and Roesch studies both found that the CL method produced fewer OTUs in the subset than in the rarefied data. In addition, the He study found that the SL method produced more OTUs, the AGC produced fewer, and the other methods produced similar numbers of OTUs than expected when comparing the subsetted data to the rarefied data. Our results support those of these previous studies (Figure 2). It was clear that inter-method differences were generally more pronounced than the differences observed between rarefying from the full dataset and from clustering the subsetted data. The number of OTUs observed was largest using the CL method, followed by the open-reference method. The AL, AGC, and DGC methods all provided comparable numbers of OTUs. Finally, the closed-reference and SL methods generated the fewest number of OTUs.

Third, the authors attempted to describe the effects of the OTU assignment instability on comparisons of communities. They used Adonis to test whether the community structure represented in subsetted communities resembled that of the full dataset when only using the unstable OTUs [@Anderson2001]. Although they were able to detect significant p-values, they appeared to be of marginal biological significance. Adonis R statistics close to zero indicate the community structures from the full and subsetted datasets overlapped while values of one indicate the communities are completely different. The He study observed adonis R statistics of 0.02 (closed-reference), 0.03 (open-reference), 0.07 (CL, AGC, DGC), and 0.16 (SL and AL). Regardless of the statistical or biological significance of these results, the analysis was tautological since, by definition, representing communities based on their unstable OTUs would yield differences. Furthermore, the *de novo* and open-reference approaches do not consistently label the OTUs that sequences belong to when the clustering methods are run multiple times with different random number seeds. To overcome this, the authors selected representative sequences from each OTU and used those representative sequences to link OTU assignments between the different sized sequence sets. It was not surprising that the only analysis that did not provide a significant p-value was for the closed-reference analysis, which is the only analysis that provides consistent OTU labels. Finally, the authors built off of this analysis to count the number of OTUs that were differentially represented between the subsetted and full datasets by each method. This entire analysis assumed that the OTUs generated using the full dataset were correct, which was an unsubstantiated assumption since the authors did not assess the quality of the OTU assignments.

@@

***How do the various methods vary in the quality of their OTU assignments?*** More important than the stability of OTUs is whether sequences are assigned to the correct OTUs. A method can generate highly stable OTUs, but the OTU assignments may be meaningless if they poorly represent the specified cutoff and the actual distance between the sequences. To assess the quality of OTU assignments by the various methods, we made use of the pairwise distance between the unique sequences to count the number of true positives and negatives and the number of false positives and negatives for each method and sampling depth. Those pairs of sequences that were similar to each other and found in the same OTU were called true positives while those that were similar and found in different OTUs were called false negatives. Meanwhile, those pairs of sequences that were different from each other and found in the same OTU were called false positives and those that were dissimilar and found in different OTUs were called true negatives. Counting the frequency of these different classes allowed us to judge how each method balanced the ratio of true positives and negatives to false positives and negatives using the MCC. We used the average MCC value as a measure of a method's quality and its variation as a measure of its consistency. We made three important observations. First, each of the *de novo* methods varied in how sensitive their MCC values were to additional sequences (Figure 1C). The SL and CL methods were the most sensitive; however, the quality of the OTU assignments did not meaningfully differ when 80 or 100% of the data were assigned to OTUs using the *de novo* methods. Second, the AL method had higher MCC values than the other methods followed by DGC, AGC, CL, open-reference, and closed-reference, and SL (Figure 1D). Third, with the possible exception of the CL method, the MCC values for each of the methods only demonstrated a small amount of variation between runs of the method with a different ordering of the input sequences. This indicates that although there may be variation between executions of the same method, they produced OTU assignments that were of equal quality. Revisiting the concept of stability, we question the value of obtaining stable OTUs when the full dataset is not optimally assigned to OTUs. Our analysis indicates that the most optimal method for assigning the Canadian soils sequences to OTUs using a 97% threshold was the AL method.

@@

***How generalizable are these results to modern datasets generated using a large number of sequences that were deeply sequenced?*** Three factors make the Canadian soil dataset less than desirable to evaluate clustering methods. First, it was one of the earliest 16S rRNA gene sequence datasets published using the 454 FLX platform. Developments in sequencing technology now permit the sequencing of millions of sequences for a study. In addition, because the original Phred quality scores and flowgram data are not available, it was not possible for us to adequately remove sequencing errors [@Schloss2011Reducing]. The large number of sequencing errors that one would expect to remain in the dataset are likely to negatively affect the performance of all of the clustering methods. Second, the dataset used in the He study covered the V9 region of the 16S rRNA gene. For a variety of reasons, this region is not well represented in databases, including the reference database used by the closed and open-reference methods. Of the `r n_v19_total` sequences in the default QIIME database, only `r n_v9_overlap` fully cover the V9 region. In contrast, `r n_v4_overlap` of the sequences fully covered the V4 region. Inadequate coverage of the V9 region would adversely affect the ability of the reference-based methods to assign sequences to OTUs. Third, our previous analysis has shown that the V9 region evolves at a rate much slower than the rest of the gene [@Schloss2010]. With these points in mind, we compared the clustering assignment for each of these methods using a time series experiment that was obtained using mouse feces [@Schloss2012Stabilization; @Kozich2013]. The MiSeq platform was used to generate `r n_miseq_seqs` sequences from the V4 region of the 16S rRNA gene of `r n_groups` samples. Parallel sequencing of a mock community indicated that the sequencing error rate was approximately 0.02% [@Kozich2013]. Although no dataset is perfect for exhaustively testing these clustering methods, this dataset was useful for demonstrating several points. First, when using 60% of the data, the stability relationships amongst the different methods were similar to what we observed using the Canadian soil dataset (Figure 3AB). With the exception for the clusters generated using CL, the methods all performed very well with stabilities greater than `r round(min(miseq_stability[miseq_stability$fraction=="0.6" & miseq_stability$method != "fn","mean"]), 2)`. Second, the MCC values calculated relative to the distances between sequences were generally higher than was observed for the Canadian soil dataset for all of the methods except the CL and SL methods. Surprisingly, the MCC values for the DGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "dgc","mean"],2)`)  and AGC (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "agc","mean"],2)`) methods were comparable to the AL method (`r round(miseq_mcc[miseq_mcc$fraction==1.0 & miseq_mcc$method == "an","mean"],2)`; Figure 3CD). This result suggests that the optimal method is likely to be database-dependent.

Finally, as was observed with the Canadian soil dataset, there was little variation in MCC values observed among the 30 randomizations. Therefore, although the methods have a stochastic component, the OTU assignments did not vary meaningfully between runs. The results from both the Canadian soil and murine microbiota datasets demonstrate that the *de novo* methods can generate stable OTU assignments and that the the overall quality of the assignments were consistent. Most importantly, these analyses demonstrate that the OTU assignments using the AL, AGC, and DGC *de novo* methods were consistently better than either of the reference-based methods.

***How does the stability and quality of OTU assignments generated by new methods compare to those analyzed in the He study?*** The Swarm algorithm is a recently proposed *de novo* method for assigning sequences to OTUs that uses user-defined parameters to break up chains generated by SL clustering [@Mah2014]. Swarm was originally validated by comparing the results against the expected clusters formed based on the taxonomic composition of a mock community. Similar to the authors of the He study, the Swarm developers suggest that methods are needed that are insensitive to input order. Use of Swarm on the Canadian soil and murine datasets demonstrated that similar to the other *de novo* methods, Swarm's OTU assignments changed as sequences were added (Figures 1A and 3A). When we compared the OTU assignments for both datasets when using all of the sequence data, the variation in MCC values across the 30 randomizations were not meaningfully different (Figures 1D and 3D). Most importantly, when we selected the distance threshold that optimized the MCC value, the quality of the OTU assignments was close to that of the AL assignments when using the Canadian soil dataset and considerably worse than that of the murine dataset (Figures 1D and 3D). Interestingly, the distance thresholds that resulted in the largest MCC values were 3 and 2% for the Canadian soil and murine datasets, respectively. This suggests that distance-based OTU definitions are not consistent across datasets when using the Swarm algorithm, although they do appear to be within the neighborhood of 3%. Finally, the Swarm developers indicated that hierarchical *de novo* algorithms were too impractical to use on large MiSeq-generated datasets. Our ability to apply AL to the large mouse dataset and even larger datasets suggests that it is not necessary to sacrifice OTU assignment quality for speed [e.g. @Schubert2015; @Zackular2015].

@@

We also observed that regardless of whether we used USEARCH or VSEARCH, the reference OTU labels that were assigned to each OTU differed between randomizations. When we used USEARCH to perform closed-reference clustering, an average of `r round(100*u_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*u_overlap["orig", "min"], 2)` to `r round(100*u_overlap["orig", "max"], 2)`%). If we instead used VSEARCH an average of `r round(100*v_overlap["orig", "mean"], 2)`% of the labels were shared between pairs of the 30 randomizations (range=`r round(100*v_overlap["orig", "min"], 2)` to `r round(100*v_overlap["orig", "max"], 2)`%). To better understand this result, we further analyzed QIIME's reference database. We hypothesized that within a given region there would be sequences that were more than 97% similar and possibly identical to each other. When a sequence was used to search the randomized databases, it would encounter a different reference sequence as the first match with each randomization. Among the `r n_v4_overlap` reference sequences that fully overlap the V4 region, there were `r n_small_v19_dist` pairs of sequences that were more than 97% similar to each other over the full length of the 16S rRNA gene. When the extracted V4 sequences were dereplicated, we identified `r n_v4_unique` unique sequences. Among these dereplicated V4 sequences there were `r n_small_v4_dist` pairs of sequences that were more than 97% similar to each other. The presence of duplicate and highly similar V4 reference sequences explains the lack of labeling stability when using either USEARCH or VSEARCH to carry out the closed-reference method. We suspect that the reference database was designed to only include sequences that were at most 97% similar to each other as a way to overcome the limitations of the USEARCH search heuristic.
